---
title: "Getting text out of files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting text out of files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_pdftools <- requireNamespace("pdftools", quietly = TRUE)
```

Before a model can read a document, the text has to come out of the file. That
step — *ingestion* — is easy to overlook and hard to undo: text that is lost here,
or mangled by an over-eager clean-up, is text no model will ever see, however
good the rest of the pipeline is. This guide covers what readgpt can read, what
it keeps, what it removes, and how to check a folder before spending anything on
it.

New to readgpt? `vignette("readgpt")` introduces the package and the ideas used
here.

```{r message = FALSE}
library(readgpt)
old <- gr_options(verbose = FALSE)
```

## What ingestion produces

`gr_ingest()` reads a file and returns a *document*: the text split into
*blocks*, which are the paragraphs, headings and list items as the file has them.
readgpt includes a short example, an annual report written in Markdown:

```{r}
doc <- gr_ingest(readgpt_example())
doc
```

Each block remembers where it came from — the page, for formats that have pages,
and the section heading it sits under:

```{r}
doc$blocks[1:6, c("block_id", "page", "section", "kind")]
```

This *provenance* is what lets an answer point back to where it came from:
`ans$evidence` reports the page and section of every chunk an answer rests on.
`page` is `NA` here because Markdown has no pages; a PDF fills it in.

`doc$text` is all the blocks joined, and `doc$stats` summarises the document:

```{r}
unlist(doc$stats[c("blocks", "chars", "tokens", "chars_removed")])
```

You rarely need to call `gr_ingest()` yourself — `answer_document()` does it for
you — but calling it directly is the way to see what the model will get.

## What readgpt can read

Each file type is handled by an *extractor*, chosen by the file's extension:

```{r}
gr_extractors()[, c("name", "extensions")]
```

| extractor | what it keeps | needs |
|---|---|---|
| `txt` | paragraphs, split at blank lines | nothing |
| `md` | paragraphs, plus headings as sections and code blocks and tables marked as such | nothing |
| `html` | text inside `<h1>`–`<h4>`, `<p>`, `<li>`, `<td>`, `<pre>` and `<blockquote>`; text anywhere else (a bare `<div>`, `<h5>`, `<th>`) is dropped, as are scripts, styles, navigation bars and footers | `xml2` |
| `docx` | paragraphs, with Word's heading styles as sections; images inside the file are read by OCR when `tesseract` is installed | `xml2` |
| `pdf` | the text of each page, with its page number; pages with no text are read by OCR | `pdftools` (and `tesseract` and `magick` for OCR) |
| `image` | the text in a picture, by OCR | `tesseract` |

A file whose extension no extractor claims — an old `.doc`, a spreadsheet — is
not read. `gr_inventory()`, below, finds those before a run rather than during
one.

You can also pass text directly. A string that is not the path of an existing
file is read as the document itself:

```{r}
memo <- gr_ingest("Revenue rose to 45.2 million dollars.\n\nHeadcount grew to 1,204.")
memo$blocks[, c("block_id", "text")]
```

The exception is a string that ends in an extension some extractor claims but
names no file. That is an error:

```{r error = TRUE}
gr_ingest("anual-report.pdf")
```

Any other string is read as text, including a path whose extension is mistyped
or is one no extractor claims:

```{r}
typo <- gr_ingest("reports/2024/annual-report.pfd")
typo$text
typo$source
```

`source` is how to tell the two apart: it holds the path when a file was read,
and `<inline text>` when the string itself was.

## Sections and headings

For Markdown, HTML and Word files, each block records the heading it falls under.
The example report has these sections:

```{r}
unique(doc$blocks$section)
```

Sections matter later. The `structural` way of cutting a document into chunks
never lets a chunk straddle two sections, and the `preview` reading strategy
plans its reading section by section. A plain text file has no headings, so its
blocks have no section.

## PDFs

A PDF is read page by page, and every block carries its page number. That is what
lets a quotation be traced to "page 12" rather than to "somewhere in the file".

```{r eval = has_pdftools}
pdf_file <- tempfile(fileext = ".pdf")
grDevices::pdf(pdf_file, width = 8.5, height = 11)
for (page in list(c("Summary", "Revenue was 45.2 million dollars in fiscal 2024."),
                  c("Outlook", "The board expects revenue of 50 million dollars in 2025."))) {
  graphics::plot.new()
  graphics::text(0, c(0.95, 0.9), page, adj = 0)
}
invisible(grDevices::dev.off())

report <- gr_ingest(pdf_file)
report$blocks[, c("page", "text")]
```

PDFs from the same source vary a lot. Text in several columns, footnotes, running
headers and tables can come out in an order that reads oddly; check a few pages of
`report$text` before relying on a new kind of PDF.

## Scanned pages and OCR

A scanned PDF is a picture of text, with no text inside it to extract. Reading it
needs *optical character recognition* (OCR), which turns the picture back into
text. OCR is slow — seconds per page — and never perfect, so readgpt uses it only
where it is needed:

```{r eval = FALSE}
# The default: OCR any page with fewer than 40 characters of text.
gr_ingest("scan.pdf", gr_ingest_spec(ocr = "auto"))

# A page with a little text but mostly scanned content: raise the threshold.
gr_ingest("scan.pdf", gr_ingest_spec(ocr = "auto", ocr_min_chars = 200))

# Everything, or nothing.
gr_ingest("scan.pdf", gr_ingest_spec(ocr = "always"))
gr_ingest("born-digital.pdf", gr_ingest_spec(ocr = "never"))

# Another language, at a higher resolution.
gr_ingest("scan.pdf", gr_ingest_spec(ocr_lang = "deu", ocr_dpi = 400))
```

OCR on PDFs needs both `tesseract` and `magick`. Without them a scanned page
comes back nearly empty and readgpt warns (`gr_ocr_unavailable`); a document with
too little text left is refused outright, as the next section shows. The
`"scanned"` recipe is set up for this kind of document.

## Cleaning

Extracted text is rarely clean. PDFs break words across lines with hyphens, repeat
the page number on every page, and use typographic ligatures ("ﬁ" for "fi").
readgpt tidies this with a series of *cleaners*, each doing one job, each of which
you can turn on or off:

```{r}
gr_cleaners()[, c("name", "stage", "default_on")]
```

The five marked `default_on` run unless you say otherwise. Note what is **off** by
default, and why:

- `remove_numbers` replaces every digit with a space, which makes any question
  about a figure, date or percentage unanswerable.
- `captions` removes lines that look like figure and table captions, which
  destroys documents made mostly of tables.
- `urls` and `emails` remove links and addresses, which are sometimes the answer.
- `references` cuts everything from a line reading just "References",
  "Bibliography" or "Works cited" to the end of the document, when that line
  is in the last 40% of the text — which loses the reference list if your
  question is about the sources. A Markdown heading such as `## References`
  is not such a line, so it does not trigger the cut.

Cleaning cannot be undone later — whatever it removes, no reader can recover — so
it is worth knowing what each choice costs.

### Presets

Instead of naming cleaners one by one, use a *preset*:

| preset | what it runs |
|---|---|
| `"none"` | nothing |
| `"minimal"` | control characters and whitespace only |
| `"standard"` | the default: page numbers, broken hyphens, control characters, ligatures, whitespace |
| `"academic"` | standard, plus `captions` and `references` |
| `"scan"` | standard, plus repeated page headers and footers |
| `"legacy"` | what the first version of this package did, number stripping included — for comparison only |

Here is how much text each leaves of the example report:

```{r}
presets <- c("none", "minimal", "standard", "academic", "scan", "legacy")
vapply(presets, function(p) gr_ingest(readgpt_example(), gr_ingest_spec(clean = p))$stats$chars,
       integer(1))
```

Or name the cleaners yourself:

```{r}
tidy <- gr_ingest(readgpt_example(),
                  gr_ingest_spec(clean = c("page_numbers", "hyphenation", "collapse_whitespace")))
tidy$stats$chars
```

### Seeing what cleaning removed

Every document records what each cleaner took out, in characters:

```{r}
vapply(doc$stats$clean_log, function(step) step$chars_removed, integer(1))
```

A negative number means a cleaner added characters — expanding "ﬁ" to "fi", say.
If a cleaner removed far more than you expected, turn it off and compare.

`gr_clean()` runs cleaners on any text, so you can try a configuration on a few of
your own lines first. Cleaners always run in two stages, `early` before `late`,
whatever order you list them in; otherwise number removal would strip the "1"
from "Page 1" before the page-number cleaner could recognise the line:

```{r}
cleaned <- gr_clean(c("Page 1", "Revenue rose to 45.2 million in 2024."),
                    steps = c("remove_numbers", "page_numbers"))
as.character(cleaned)
```

The second line shows why `remove_numbers` is off by default.

## When too little text is left

A document with almost no text after cleaning — a scan read without OCR, an empty
file, a PDF of images — is refused rather than sent to the model, because a model
shown nothing would answer "not in the document" for a document that may well
contain the answer:

```{r error = TRUE}
gr_ingest("Tiny.")
```

The limit is `min_chars` (20 characters by default) in `gr_ingest_spec()`.

## Reading the same file twice

Ingestion is cached for the rest of the R session. Asking a second question of the
same file, or trying a different way of cutting it into chunks, reuses the
extracted text instead of reading the file again. The cache notices when the file
changes (its size or modification time) and when any ingestion setting changes.
Turn it off with `gr_options(cache_documents = FALSE)`.

## Before reading a whole folder

A folder of documents hides problems that a single file makes obvious: a few old
`.doc` files nobody can read, scanned PDFs without text, an empty file, or a
subfolder you forgot about. `gr_inventory()` finds all of these in seconds,
without calling a model and without an API key:

```{r}
inbox <- file.path(tempdir(), "inbox")
dir.create(inbox, showWarnings = FALSE)
invisible(file.copy(readgpt_example(), file.path(inbox, "annual_report.md"),
                    overwrite = TRUE))
writeLines("Headcount grew to 1,204 employees across nine sites.",
           file.path(inbox, "notes.txt"))
writeLines("(a Word 97 document)", file.path(inbox, "minutes.doc"))
invisible(file.create(file.path(inbox, "empty.txt")))

inv <- gr_inventory(inbox)
inv
```

It lists every file, including the ones that will not be read, because "one file
will be skipped" is exactly what you need to know before the run:

```{r}
inv$files[, c("file", "extractor", "status", "tokens")]
```

The statuses:

| status | meaning |
|---|---|
| `ready` | an extractor can read it, and there is text |
| `needs_ocr` | a PDF whose pages have no text; it needs OCR |
| `needs_package` | an extractor can read it, but its package is not installed |
| `no_extractor` | no extractor reads this kind of file |
| `empty` | nothing to read |
| `unreadable` | it could not be opened; `note` says why |

`inv$totals` adds up the tokens and prices one request per readable file with
your model: its tokens in, and an assumed 500-token answer out. Treat
`cost_floor_usd` as a rough guide rather than a bound — reading strategies that
make several requests per document cost more, and short answers cost less.
Files whose size is unknown until they are read by OCR are not in the total;
`tokens_unknown` counts them.

```{r}
unlist(inv$totals[c("readable", "tokens", "cost_floor_usd")])
```

## Adding your own

A format readgpt does not know, or a clean-up step your documents need, can be
added and then used exactly like a built-in one. A cleaner that drops lines
marked "CONFIDENTIAL":

```{r}
gr_register_cleaner("drop_confidential", stage = "early",
  fn = function(x, opts) gsub("(?mi)^\\s*CONFIDENTIAL.*$", "", x, perl = TRUE),
  description = "Remove lines marked CONFIDENTIAL")

gr_ingest("CONFIDENTIAL - internal only\n\nRevenue rose to 45.2 million dollars.",
          gr_ingest_spec(clean = c("drop_confidential", "collapse_whitespace")))$text
```

`?gr_register_extractor` shows how to add a file format.

```{r include = FALSE}
gr_options(old)
```

## Next

Once the text is out, the next decision is how to cut it into chunks and how the
model should read them. `vignette("readers")` covers the reading strategies;
`vignette("tour")` shows the chunking methods.
