---
title: "Reading a document with a language model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading a document with a language model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Asking a language model a question about a document involves three decisions,
and they are usually made for you, invisibly, all at once. This package makes
them separate:

- **ingest** — get text out of the file, and decide what to throw away
- **segment** — cut that text into pieces small enough to send
- **read** — decide how the model works through those pieces to an answer

Each is a registry you can list, swap and extend. Any ingest × any segmenter ×
any reader composes.

```{r message = FALSE}
library(readgpt)
```

## Running this vignette without an API key

Everything below runs offline. Two settings make that work, and they are worth
understanding because they are also how you develop against this package without
spending anything.

```{r}
cl <- gr_mock_client(function(messages, params) "Revenue was 45.2 million dollars.")
old <- gr_options(verbose = FALSE, embedder = "lexical")
```

`gr_mock_client()` is a client whose handler is an ordinary R function. It goes
everywhere a real client goes and records every prompt it was sent, which is
what lets you check that a reading strategy is doing what you think before you
pay for it.

`embedder = "lexical"` selects hashed bag-of-words vectors instead of an
embeddings endpoint. They measure word overlap rather than meaning, so they are
a poor substitute in production — but they are free, offline, and *deterministic*,
which is what makes this document reproducible. `gr_embedders()` lists what is
registered.

## A first run

`answer_document()` binds one ingest, one segmenter and one reader into a
pipeline and runs it. The bundled example is a short annual report.

```{r}
ans <- answer_document(readgpt_example(), "What was revenue?", "thorough",
                       client = cl)
ans$answer
```

`"thorough"` names a recipe. `gr_recipes()` returns the built-ins, and each is
just a binding of the three axes:

```{r}
do.call(rbind, lapply(names(gr_recipes()), function(n) {
  r <- gr_recipes(n)
  data.frame(recipe = n, segment = r$segment$method, reader = r$read$reader)
}))
```

`gr_recipes("thorough")` prints one in full, showing every setting on all three
axes. A recipe is only a starting point: any of its settings can be overridden
in the call, and `gr_recipe()` builds one from scratch.

## Reading the answer

The answer is not a string. It is an object that says how it was arrived at, and
the first thing to look at is never `$answer`:

```{r}
ans$partial
```

`partial` is `TRUE` whenever anything degraded — a call failed, chunks were
dropped for space, a budget stopped the run early, an embedding fell back to
lexical vectors. **Check it before trusting the answer.** When it is `TRUE`,
`$notes` says what happened.

```{r}
names(ans$notes)
```

`$evidence` is what the answer rests on. What `text` holds depends on the
reader: verbatim chunk text for `stuff`, `retrieve`, `rerank` and `iterative`;
model-extracted passages for `skim`; per-chunk model answers for `map_reduce`.

```{r}
ans$evidence[, c("chunk_id", "page", "section")]
```

And `$trace` records every prompt, response and token count from that one run —
not from a second run made to explain the first.

```{r}
gr_trace_summary(ans$trace)[, c("calls", "cached", "tokens_in", "tokens_out")]
```

## Did the model quote the document, or invent the quote?

For most readers `$evidence` is verbatim chunk text and is true by construction:
the package put it there. For `skim` it is not. `skim` asks the model to extract
the passages that bear on the question, and what comes back is whatever the model
chose to write — presented as a quotation, with nothing checking that it was one.

That is the worst gap to leave. A fabricated citation is more convincing than a
fabricated answer, because it looks like the thing that would let you check.

```{r}
doc <- "Revenue rose to 45.2 million dollars.\n\nHeadcount grew to 1,204."
ch  <- gr_segment(gr_ingest(doc), list(method = "paragraph", max_tokens = 40))

extracting <- function(quote) gr_mock_client(function(messages, params) {
  if (grepl("You extract evidence", messages[[1]]$content, fixed = TRUE)) return(quote)
  "Revenue was 45.2 million dollars."
})

faithful <- gr_read(ch, "What was revenue?",
                    extracting("Revenue rose to 45.2 million dollars."), "skim")
gr_verify_evidence(faithful)[, c("kind", "verified", "match")]
```

Now a model that invents one. The span is fluent, on topic, attributed to a real
chunk, and not in the document — the case you cannot catch by eye:

```{r}
invented <- gr_read(ch, "What was revenue?",
                    extracting("Revenue rose to 88.9 billion dollars on record demand."),
                    "skim")
gr_verify_evidence(invented)[, c("kind", "verified", "match")]
invented$partial
invented$notes$unverified_evidence
```

The comparison is forgiving about typography and unforgiving about content.
Whitespace, curly quotes, dashes, case and the punctuation a model wraps a quote
in are all folded away, because none of that is fabrication and flagging it would
make `partial` stop meaning anything. A changed number is not folded away. Below
an exact match, `match` is the fraction of the span carried by its longest
consecutive run in the source — a run rather than word overlap, because overlap
cannot tell a quotation from a paraphrase built from the same words.

Citations get the same treatment, for every reader. An answer citing a chunk that
was never sent to it sets `notes$cited_unknown` and is `partial`.

## Axis 2: segmentation is free to experiment with

Chunking decides what the model can possibly see together. It involves no model
calls at all, so you can compare strategies for nothing before spending anything
on reading:

```{r}
doc <- gr_ingest(readgpt_example())
do.call(rbind, lapply(c("fixed", "paragraph", "sentence", "structural"), function(m)
  gr_chunk_stats(gr_segment(doc, list(method = m, max_tokens = 120)))))
```

`over_cap` is the column to watch: a chunk over the cap is one the reader will
have to truncate or drop. `gr_segmenters()` lists all nine, with which of them
need a client (the `semantic` one embeds).

Overlap matters too. Without it, an answer straddling a boundary is lost by both
chunks:

```{r}
do.call(rbind, lapply(c(0, 40), function(ov)
  gr_chunk_stats(gr_segment(doc, list(method = "sentence", max_tokens = 120,
                                      overlap_tokens = ov)))))
```

## Axis 3: the readers are genuinely different

Nine strategies, each declaring a *traversal signature* — how it selects chunks,
how many calls it makes, and what state it carries between them:

```{r}
gr_readers()[, c("name", "signature", "cost_calls")]
```

The signature is not documentation, it is a checkable property. Two recipes that
resolve to the same segmentation and the same signature are the same work, and
`gr_compare()` refuses to bill you twice for them.

```{r}
cmp <- gr_compare(readgpt_example(), "What was revenue?",
                  c("fast", "precise", "thorough"), client = cl)
cmp$summary[, c("recipe", "segmenter", "chunks", "reader", "chunks_used", "not_found")]
```

Extraction is shared across recipes and segmentation is shared between recipes
whose segment specs match, so comparing three readers over one chunking costs one
chunking. One trace covers the whole comparison:

```{r}
gr_trace_summary(cmp$trace)[, c("calls", "cached", "tokens_in")]
```

`not_found` distinguishes "the document does not say" from a failure. So does
`is_not_found()` on any answer. A model that invents an answer is a worse outcome
than one that admits the document is silent, and the two must not look alike.

## Choosing which chunks, and where to put them

Top-k by similarity answers "which chunks are most like the question", which is
not quite the question you wanted answered. If three paragraphs say the same
thing, all three score highly and all three go in the prompt.

```{r}
redundant <- paste(c(
  "Revenue was 45.2 million dollars in fiscal 2024.",
  "Total revenue reached 45.2 million dollars in the 2024 fiscal year.",
  "In fiscal 2024 the company recorded revenue of 45.2 million dollars.",
  "Headcount grew to 1,204 employees across nine clinical sites.",
  "The board approved a dividend of 0.42 dollars per share in March."),
  collapse = "\n\n")

ch <- gr_segment(gr_ingest(redundant), list(method = "paragraph", max_tokens = 40))
picked <- function(m) {
  gr_read(ch, "What was revenue?", cl,
          list(reader = "retrieve", top_k = 3, mmr = m))$chunks_used
}
rbind("mmr = 1 (top-k)" = picked(1), "mmr = 0.3" = picked(0.3))
```

`mmr` below 1 trades relevance against redundancy, so the second and third slots
go to chunks that add something. It costs nothing — the vectors are already
computed.

`context_order` is a separate decision about *placement*: transformers attend
better to the beginning and end of a long context than to its middle, so
`"edges"` puts the strongest chunk first and the second-strongest last. It never
changes which chunks were selected, only where they sit.

## Rails, before you spend anything

Two caps are on by default and checked *before* the first request:

```{r}
unlist(gr_options()[c("max_cost_usd", "max_calls")])
```

You can also ask what a run would cost before making it:

```{r}
gr_estimate_cost("gpt-4o", input_tokens = 120000, output_tokens = 4000)
```

And what one did cost afterwards — counting only the calls actually issued,
which is not the same as the tokens it moved:

```{r}
gr_trace_cost(ans$trace)[, c("model", "calls", "paid_calls", "usd")]
```

## Making a re-run free, and a result checkable

Every stage of this package except the model call is a pure function of its
input. The model call is the only step that costs money, the only one that can
die halfway through a long run, and above a temperature of zero the only one that
does not return the same thing twice. Two features follow from that.

A **cache** stores each successful response against the exact request, so a
repeat is free and identical:

```{r}
cache <- gr_cache(dir = file.path(tempdir(), "readgpt-vignette-cache"))
cached_cl <- gr_cache_client(cl, cache)

first  <- answer_document(readgpt_example(), "What was revenue?", "fast",
                          client = cached_cl)
second <- answer_document(readgpt_example(), "What was revenue?", "fast",
                          client = cached_cl)

gr_trace_summary(second$trace)[, c("calls", "cached")]
```

`calls - cached` is what a run actually paid for. Failures are never cached: a
rate limit is a property of the moment, and storing one would make a blip
permanent.

A **replay** goes further. A trace already holds every prompt and every response,
so with the document it is enough to reproduce the run exactly — no key, no
network, no spend:

```{r}
f <- file.path(tempdir(), "run.json")
gr_trace_save(first$trace, f)

replayed <- answer_document(readgpt_example(), "What was revenue?", "fast",
                            client = gr_replay_client(f))
identical(replayed$answer, first$answer)
```

That is the difference between a result someone has to trust and one they can
check. A prompt with no recorded response raises `gr_replay_miss` rather than
inventing an answer, because a result that looks like the original and is not is
worse than no replay at all.

Embeddings are not model calls and are not in the trace, so a replay reproduces
chunk *ranking* only when the recording used a deterministic embedder and the
replay uses the same one — which is why this vignette set `embedder = "lexical"`
at the top.

## Many documents

`gr_compare()` runs several recipes over one document. `gr_read_many()` runs one
recipe over many, and gives you a row per document:

```{r}
folder <- file.path(tempdir(), "reports")
dir.create(folder, showWarnings = FALSE)
writeLines("Revenue was 45.2 million dollars in fiscal 2024.",
           file.path(folder, "north.txt"))
writeLines("Revenue was 51.8 million dollars in fiscal 2025.",
           file.path(folder, "south.txt"))

out <- gr_read_many(folder, "What was revenue?", "fast", client = cl)
out$summary[, c("document", "not_found", "chunks_used", "calls", "status")]
```

One unreadable file is one `"failed"` row, not a dead run. Each document gets its
own budget, so one enormous file cannot starve the rest. `store =` makes the run
resumable, and `max_total_usd` is a corpus-wide ceiling.

## From a folder to a review

The three axes answer a question. A corpus job usually wants a *table*, and a
review wants a table plus the account of it. Four functions cover that, and they
compose in one direction:

`gr_protocol()` → `gr_screen()` → `gr_extract()` → `gr_synthesise()`

A protocol is what you fix before reading anything: which documents count, what
to collect from the ones that do, and what the write-up has to cover. Fixing it
first is the point — a criterion invented while reading is a criterion fitted to
what was found.

```{r}
protocol <- gr_protocol(
  "revenue-review",
  question = "How did revenue change across the regional reports?",
  include  = "Reports a revenue figure",
  exclude  = "Is a forecast rather than a result",
  fields   = gr_fields(
    region  = "The region the report covers",
    revenue = gr_field("Revenue in millions of dollars", type = "number"),
    year    = gr_field("Fiscal year reported", type = "integer")
  ),
  outline  = c("Findings" = "How revenue compares across regions")
)
protocol
```

`gr_protocols()` lists three templates to start from, and
`gr_protocol_save()`/`gr_protocol_read()` round-trip one through a JSON file so
it can be shared and cited alongside the results.

Screening is one call per document. Every document gets a decision and a reason,
and nothing is dropped on the way — a file that could not be read has no
decision rather than a silent exclusion, and `"unclear"` is an answer rather
than a forced guess.

```{r}
screener <- gr_mock_client(function(messages, params) {
  seen <- paste(vapply(messages, function(m) as.character(m$content), character(1)),
                collapse = " ")
  line <- regmatches(seen, regexpr("Revenue was [0-9.]+ million dollars in fiscal [0-9]+\\.", seen))
  sprintf('{"decision":"include","reason":"Reports a revenue figure.",
            "criterion":"Reports a revenue figure","quote":"%s"}', line)
})

screened <- gr_screen(folder, protocol, client = screener)
screened$table[, c("document", "decision", "criterion", "verified")]
```

Extraction fills the schema from every chunk of each included document, then
reconciles. The result is one typed row per document, and every filled cell
carries the sentence it came from, checked against the chunk it was attributed
to.

```{r}
extractor <- gr_mock_client(function(messages, params) {
  seen <- paste(vapply(messages, function(m) as.character(m$content), character(1)),
                collapse = " ")
  n <- if (grepl("45.2", seen, fixed = TRUE)) "45.2" else "51.8"
  yr <- if (grepl("2024", seen, fixed = TRUE)) 2024 else 2025
  rg <- if (grepl("45.2", seen, fixed = TRUE)) "north" else "south"
  line <- sprintf("Revenue was %s million dollars in fiscal %d.", n, yr)
  sprintf('{"region":"%s","revenue":%s,"year":%d,
            "region__quote":"%s","revenue__quote":"%s","year__quote":"%s"}',
          rg, n, yr, line, line, line)
})

table <- gr_extract(screened$included, protocol, client = extractor, recipe = "fast")
table$table[, c("document", "region", "revenue", "year", "n_unverified")]
```

`n_unverified` is the column to look at before believing a row: zero means every
value in it can be pointed at in the document. `table$evidence` is the long form
— one row per supported cell, with the quote, the page it is on, and whether the
quote really appears there.

```{r}
table$evidence[, c("document_id", "field", "quote", "verified")]
```

Finally the write-up, one call per section of the outline, citing the rows it
rests on:

```{r}
writer <- gr_mock_client(function(messages, params) {
  "Revenue was higher in the southern region [study 2] than the northern [study 1]."
})

review <- gr_synthesise(table, protocol, client = writer)
review$citations
cat(review$text)
```

Every `[study n]` is parsed back out and checked against the rows that exist;
one pointing at a row that is not there is reported and marks the section
partial. That completes the chain — a sentence cites a study, the study's row
cites a quote, the quote was checked against the page it is attributed to. None
of it proves the sentence is true. It makes every step of the way back to the
document short enough to walk.

## Using a real model

Everything above used a mock. In production you need a client:

```{r eval = FALSE}
# An OpenAI-compatible endpoint:
cl <- gr_client(model = "gpt-4o")          # key from OPENAI_API_KEY

# Or any provider ellmer speaks to -- Anthropic, Google, Bedrock, Ollama:
cl <- gr_ellmer_client(ellmer::chat_anthropic())

# Or anything at all:
cl <- gr_backend_client(function(messages, params) my_provider(messages))
```

Without a key nothing raises. Every call fails, the answer is the
`"NOT_IN_DOCUMENT"` sentinel and `ans$partial` is `TRUE` — which is why checking
`partial` is the habit worth forming.

## Extending it

Every axis is a registry, and an addition behaves exactly like a built-in — same
token caps, same provenance, same reporting, and it can be named in a recipe or
put in an `ensemble`.

```{r}
gr_register_segmenter("by_bullet", description = "one chunk per bullet",
  fn = function(doc, spec, client, trace) {
    units <- unlist(strsplit(doc$text, "\n(?=[-*])", perl = TRUE))
    new_chunks(units, "by_bullet", spec)
  })

"by_bullet" %in% gr_segmenters()$name
```

`gr_register_extractor()`, `gr_register_cleaner()`, `gr_register_reader()`,
`gr_register_embedder()` and `gr_register_model()` work the same way. Each has a
worked example in its help page.

```{r include = FALSE}
gr_options(old)
```
