---
title: "Get started with readgpt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with readgpt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

readgpt asks a large language model questions about your documents (a report, a
contract, a folder of research papers) and returns the answer with a record of
how it was reached: which parts of the document the model saw, what it was
asked, what it said, and what that cost. This guide assumes no previous
experience with language models. It explains the few ideas you need, installs
the package, and goes through a first question from start to finish.

Everything below runs without an account or an internet connection, so you can
follow along for free.

## A few ideas first

**The model runs somewhere else.** A large language model (an *LLM*) such as
GPT is a program run by a *provider*: OpenAI, Anthropic, Google, or a model on
your own computer. readgpt sends it text over the internet and reads the reply.
Commercial providers charge for each request.

**Text is measured in tokens.** Models do not read characters or words; they
read *tokens*, pieces of words. In English a token is roughly four characters,
so a page of text is several hundred tokens. Providers charge per token, both
for the text you send and, at a higher rate, for the text the model writes back.

```{r message = FALSE}
library(readgpt)
```

```{r}
gr_count_tokens("Revenue rose to 45.2 million dollars in fiscal 2024.")
```

readgpt estimates token counts itself, without asking the provider, so it can
work out what a request will cost before sending it.

**A model can only take in so much at once.** The most text a model accepts in a
single request is its *context window*. A long report may not fit, and even when
it does, models tend to overlook details in the middle of a very long prompt. So
documents are cut into smaller pieces, called *chunks*, and the question is put
to the chunks in some order.

**That makes three decisions**, and readgpt keeps them separate so you can see and
change each one:

1. **Ingest**: get the text out of the file and tidy it.
2. **Segment**: cut the text into chunks.
3. **Read**: decide how the model works through the chunks. It can take all of
   them in one request, each one separately, only the most relevant few, and so
   on.

A fourth choice, which model answers, is independent of all three.

**Models can make things up.** A model may state something the document never
says, and it can do so fluently. readgpt cannot prevent that. It records which
text each answer rests on, checks quotations against the document, and tells you
when anything went wrong. The first example below shows where to look.

## Installing

readgpt needs R 4.1 or later. Install it from GitHub:

```{r eval = FALSE}
install.packages(c("remotes", "knitr", "rmarkdown"))
remotes::install_github("elkronos/readgpt", build_vignettes = TRUE)
```

`build_vignettes = TRUE` also installs these guides, so `vignette()` can open
them. Building them needs `knitr`, `rmarkdown` and Pandoc. Pandoc comes with
RStudio; without RStudio, install it from <https://pandoc.org/installing.html>
first, or leave out `build_vignettes = TRUE`.

The package itself depends only on `digest`, `httr` and `jsonlite`. Some file
types and features need an extra package. Each is optional, so install only what
you use:

| to do this | install |
|---|---|
| read PDFs | `pdftools` |
| read Word documents and web pages | `xml2` |
| read scanned pages and images (OCR) | `tesseract` and `magick` |
| use a provider other than OpenAI | `ellmer` |
| run many requests at once | `future` and `future.apply` |
| count tokens exactly as OpenAI does | `reticulate`, with Python's `tiktoken` |
| use the point-and-click app | `shiny` |

If you ask for something that needs a package you do not have, readgpt says
which one. It stops, unless it can carry on without it. For example, it runs
requests one at a time when the parallel packages are missing.

## Connecting to a model

To use a real model you need an account with a provider and an *API key*, a
password that identifies your account to the provider's servers. For OpenAI,
create one at <https://platform.openai.com>.

Keep the key out of your scripts. The usual place is your `.Renviron` file, which
R reads when it starts:

```{r eval = FALSE}
file.edit("~/.Renviron")
# add this line, save the file, then restart R:
# OPENAI_API_KEY=sk-...
```

`gr_api_key()` checks that R can see it, and stops with an explanation if it
cannot:

```{r eval = FALSE}
gr_api_key()
```

The README's "Other providers" section covers Anthropic, Google, Azure, company
gateways, and free models running on your own machine through Ollama.

## Practising without a key

For learning, a real model is unnecessary. `gr_mock_client()` builds a stand-in
that replies with whatever an R function returns, and it can be used anywhere a
real model can. This one always gives the same answer, which is enough to see
how everything else works:

```{r}
cl <- gr_mock_client(function(messages, params) "Revenue was 45.2 million dollars.")
old <- gr_options(verbose = FALSE)
```

`verbose = FALSE` stops readgpt printing a line for each stage; leave it on in
your own work while you are learning. `gr_options()` returns the settings it
replaced, so they can be restored later with `gr_options(old)`.

## Your first question

`answer_document()` does all three steps in one call. readgpt includes a short
example document, an annual report:

```{r}
ans <- answer_document(readgpt_example(), "What was revenue in 2024?", client = cl)
ans$answer
```

With a real model you would leave out `client =` and point it at your own file:

```{r eval = FALSE}
ans <- answer_document("annual-report.pdf", "What was revenue in 2024?")
```

The first argument can be a path to a file or the text itself. A missing file
whose name ends in an extension readgpt reads (`.pdf`, `.docx`, `.md` and so on)
is an error. Any other string is read as the text of the document, including a
path with a mistyped extension: `"reports/2024/annual-report.pfd"` becomes a
one-line document about a file name. `ans$document$source` shows which happened:
the path for a file, `<inline text>` for a string.

## Checking the answer

`ans` holds much more than the answer. Look at three things before relying on
it.

**`partial` says whether anything went wrong.** It is `TRUE` if a request
failed, if chunks were left out because they did not fit, if the limit on
requests stopped the run early, or if a reading strategy fell back to a simpler
method. Some problems only raise a warning, such as a scanned page that could
not be read, so read any warnings as well:

```{r}
ans$partial
```

When it is `TRUE`, `ans$notes` says why. Here is a "model" that fails every
request, as a network outage would:

```{r}
broken <- gr_mock_client(function(messages, params) stop("no connection"))
bad <- answer_document(readgpt_example(), "What was revenue in 2024?",
                       client = broken)
bad$partial
names(bad$notes)
```

The run did not stop with an error. It finished, and the failure is recorded on
the result, so check `partial` every time.

**`is_not_found()` says whether the document had the answer at all.** Every
reading strategy tells the model to reply with a fixed marker when the text it
was shown does not answer the question, so "the document does not say" can be
told apart from an answer:

```{r}
is_not_found(ans$answer)
is_not_found(bad$answer)
```

For `bad`, the marker means "no answer came back", and `partial` says that was a
failure rather than a finding.

**`evidence` shows what the answer rests on.** Each row is a chunk the answer
drew on, with where it came from: the page, for a PDF, and the heading it sits
under. This short report fitted in a single chunk spanning several headings, so
there is one row and both are empty:

```{r}
ans$evidence[, c("chunk_id", "page", "section")]
```

On a long document there is a row for every chunk used, and `page` and `section`
tell you where to find the passage. What each row's `text` holds depends on the
reading strategy. When the model is asked to quote the document, readgpt checks
every quotation against the text and marks the answer `partial` if one is not
there. `vignette("readers")` shows both.

## What it cost

Every result carries a *trace*: each request sent, each reply, and the tokens
both used.

```{r}
gr_trace_summary(ans$trace)[, c("calls", "tokens_in", "tokens_out")]
gr_trace_cost(ans$trace)[, c("model", "calls", "usd")]
```

`usd` is priced from readgpt's list of known models (`gr_models()`). A model it
does not know is priced as unknown rather than free. You can also estimate
before sending anything. Here is a 120,000-token document with a 4,000-token
answer:

```{r}
gr_estimate_cost(gr_options("model"), input_tokens = 120000, output_tokens = 4000)
```

Two safety limits are on from the start. Before the reading step begins, readgpt
prices the worst case, with every reply at its maximum length, and refuses a run
whose worst case is above $5. That figure is usually far above what the run
costs. readgpt also never makes more than 400 requests for one document:

```{r}
unlist(gr_options()[c("max_cost_usd", "max_calls")])
```

Change them with `gr_options()`, for example `gr_options(max_cost_usd = 20)`.

## Choosing a recipe

`answer_document()` made the three decisions using a *recipe*, a named
combination of choices. The default is `"thorough"`. The built-in recipes:

```{r}
do.call(rbind, lapply(names(gr_recipes()), function(n) {
  r <- gr_recipes(n)
  data.frame(recipe = n, chunks = r$segment$method, reading = r$read$reader)
}))
```

A starting point for each kind of job:

| your document and question | try |
|---|---|
| short enough to send whole | `"fast"` |
| one fact somewhere in a long document | `"needle"` |
| every mention has to be found | `"thorough"` |
| long, with headings, and you want a summary | `"survey"` |
| scanned pages, forms or invoices | `"scanned"` |
| an argument that builds through the text | `"narrative"` |
| a question that needs several facts combined | `"research"` |
| high stakes, and you want a second opinion | `"consensus"` |

Pass the name as the third argument:

```{r}
fast <- answer_document(readgpt_example(), "What was revenue in 2024?", "fast",
                        client = cl)
fast$reader
```

Recipes differ a great deal in cost. `"thorough"` sends every chunk to the model
separately, so a long document means many requests; `"fast"` and `"needle"` make
one. `vignette("readers")` explains each reading strategy and when to use it.

## More than one document

For a folder of documents, `gr_read_many()` asks the same question of each file
and returns one row per document. For a literature review, `gr_screen()` decides
which documents meet your criteria, `gr_extract()` builds a table of the details
you ask for, and `gr_synthesise()` writes up the findings with each statement
citing the study it came from. `vignette("tour")` shows each of these briefly.

```{r include = FALSE}
gr_options(old)
```

## Where to go next

- `vignette("ingest")`: getting text out of PDFs, Word files, web pages and
  scans, and cleaning it.
- `vignette("readers")`: the reading strategies, what each costs, and how to
  choose.
- `vignette("tour")`: everything else, briefly. It covers chunking, caching and
  replaying runs, many documents, reviews, and extending the package.
- `?gr_options`: every setting, with its default.

`browseVignettes("readgpt")` lists all the guides.

## Glossary

**API key**: a password identifying your account to a provider.

**Chunk**: a piece of a document small enough to send to the model.

**Context window**: the most text a model accepts in one request.

**Embedding**: a list of numbers standing for a piece of text, used to find
chunks similar to a question.

**Evidence**: the text an answer rests on, in `ans$evidence`.

**LLM**: large language model, the program that reads and writes text.

**Mock client**: a stand-in for a model that runs an R function, for practising
and testing without cost.

**Partial**: `ans$partial` is `TRUE` when a request failed, chunks did not fit
and were left out, the run stopped early, or a reading strategy fell back to a
simpler method. Some problems only raise a warning.

**Prompt**: the text sent to a model in one request.

**Provider**: the company or program that runs the model.

**Reader**: a strategy for working through the chunks.

**Recipe**: a named combination of an ingest, a segmentation and a reader.

**Segmenter**: a method of cutting text into chunks.

**Token**: the unit models read and providers charge by, roughly four characters
of English.

**Trace**: the record of every request and reply in a run.
