---
title: "Choosing how the model reads"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing how the model reads}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Once a document is cut into chunks, something has to decide how the model works
through them. Send everything at once? Ask each chunk separately and combine the
answers? Find the few chunks most like the question and ignore the rest? These
are *reading strategies*, which readgpt calls *readers*. They decide most of the
cost of a run and much of the quality of its answer. The same chunks read two
ways can take one request or dozens, and can find a fact or miss it.

This guide explains each reader in plain terms, shows how they differ on the same
document, and gives a way to choose. New to readgpt? Start with
`vignette("readgpt")`.

## Setting up

The examples use a stand-in for a real model, so they run for free. This one
answers each kind of request the way a model would: it scores an excerpt when
asked to score and quotes when asked to quote, so every reader can do its normal
work. You do not need to follow its details; it plays the part of the model.

```{r message = FALSE}
library(readgpt)
old <- gr_options(verbose = FALSE, embedder = "lexical")

stand_in <- function(messages, params) {
  instructions <- messages[[1]]$content
  everything <- paste(vapply(messages, function(m) m$content, character(1)),
                      collapse = "\n")
  task <- if (is.null(params$schema_name)) "" else params$schema_name
  if (task == "relevance") {        # rerank: score one excerpt from 0 to 10
    excerpt <- messages[[length(messages)]]$content
    if (grepl("revenue", excerpt, ignore.case = TRUE)) {
      return('{"score": 9, "reason": "States revenue."}')
    }
    return('{"score": 2, "reason": "About something else."}')
  }
  if (task == "iterative_step") {   # iterative: can I answer yet?
    return('{"can_answer": true, "answer": "Revenue was 45.2 million dollars.", "next_query": ""}')
  }
  if (task == "reading_plan") {     # preview: which sections to read
    return('{"sections": [{"id": 1, "treatment": "read", "reason": "States revenue."}]}')
  }
  if (grepl("You extract evidence", instructions, fixed = TRUE)) {   # skim: quote
    quote <- regmatches(everything,
                        regexpr("[^.\n]*revenue of [0-9.]+ million dollars[^.\n]*", everything))
    return(if (length(quote)) quote else "NONE")
  }
  "Revenue was 45.2 million dollars."
}
```

`embedder = "lexical"` makes the readers that search for similar text use simple
word matching instead of a paid embedding service; more on that below.

Now the document, cut into chunks along its section headings:

```{r}
chunks <- gr_segment(gr_ingest(readgpt_example()),
                     list(method = "structural", max_tokens = 120))
gr_chunk_stats(chunks)
```

Eight chunks. A real report would have hundreds, and the differences below would
be correspondingly larger.

## Listing the readers

```{r}
gr_readers()[, c("name", "signature", "cost_calls")]
```

`cost_calls` is how many requests a reader makes, as a formula in **N**, the
number of chunks. A reader whose count grows with N makes a hundred times as
many requests on a document a hundred times longer. But you pay for tokens, not
requests, and a reader that looks at every chunk sends every token: `stuff`
makes one request whatever the length, yet that request carries the whole
document, so its cost grows with the document too. Each request also repeats
the instructions and writes its own reply, so among readers that see everything,
more requests cost more. The readers that look at a selection of chunks (the
`topk` ones below) cost roughly the same however long the document is.

`signature` describes how a reader works, in three parts separated by `|`:

- **which chunks it looks at**: `all` of them; the `topk` most relevant; the
  `head` (opening) of the document; the chunks a `planned` reading chose; or, for
  an `ensemble`, whatever its members looked at;
- **how many requests it makes**, the same formula in short;
- **what carries over between requests**: `none` (each request stands alone),
  `tree` (answers are combined in stages), or `forward` (each request builds on
  the one before).

The built-in readers all have different signatures, and `ensemble` refuses
members that share one: two readers with the same signature would be the same
method under two names.

## Every reader on one document

Here is the same question put to the same eight chunks by each general-purpose
reader. `extract` and `screen` are left out: they are built for literature
reviews and need a schema or criteria.

```{r}
read_with <- function(reader) {
  model <- gr_mock_client(stand_in)
  ans <- gr_read(chunks, "What was revenue in 2024?", model, reader)
  data.frame(reader = reader, requests = length(model$calls()),
             chunks_used = length(ans$chunks_used), partial = ans$partial)
}
readers <- c("stuff", "map_reduce", "refine", "skim", "hierarchical",
             "retrieve", "rerank", "iterative", "preview", "ensemble")
do.call(rbind, lapply(readers, read_with))
```

The stand-in gives every reader the same answer, and none of the runs degraded
(`partial` is `FALSE`). What differs is the work: from one request to eleven for
eight chunks, and from reading everything to relying on a single passage. On a
long document the gap widens: the number of requests grows with the document for
some readers and not at all for others, and the cost grows for every reader that
reads the whole document.

## The readers

### `stuff`: everything in one request

Puts the whole document into a single prompt. It is the simplest reader, and
usually the cheapest way to have the model read all of a document, with one set
of instructions and one reply. Its cost still grows with the document's length.
The model sees everything at once, so it can connect facts from different
places. The limit is the context window: if the document does not fit,
`stuff` sends what fits, warns, and marks the answer partial
(`on_overflow = "error"` makes it stop instead). Very long prompts also dilute
attention; details in the middle of a long prompt are more easily missed.

**Use it for** documents comfortably inside the model's context window.
**Recipe:** `"fast"`.

### `map_reduce`: every chunk separately, then combined

Asks the question of each chunk on its own ("map"), then combines the answers in
stages ("reduce"). Nothing is skipped, each request is small, and the per-chunk
requests can run in parallel. It costs N requests plus the combining steps, and a
fact that only makes sense across two chunks can be missed, because no single
request sees both.

**Use it for** questions where every mention matters: "list every risk named in
this report". **Recipe:** `"thorough"`, which `answer_document()` uses by
default.

### `refine`: a draft revised chunk by chunk

Reads the chunks in order, keeping a running answer and revising it with each new
chunk. Later text can correct earlier conclusions, which suits documents whose
argument develops. It costs N requests that cannot run in parallel, and the
running answer can drift.

**Use it for** narratives and arguments that build. **Recipe:** `"narrative"`.

### `skim`: quote first, then answer

For each chunk, asks the model to copy out any passages that bear on the
question, then answers once from those passages alone. The final answer rests on
quotations, which readgpt checks against the text (see below). It costs N + 1
requests; the per-chunk step can use a cheaper model through `skim_model`.

**Use it for** questions where you want the answer tied to exact wording.
**Recipe:** `"precise"`.

### `hierarchical`: summaries of summaries

Summarises each chunk, then summarises groups of summaries (`fan_in` at a time),
level by level, until everything fits in one prompt, and answers from that. It
suits questions about a whole long document, such as "what is this report's
overall argument?", at the price of detail lost in summarising. `summary_model`
can use a cheaper model for the summaries; `max_levels` caps the depth.

**Use it for** long, structured documents where you want an overview.
**Recipe:** `"survey"`.

### `retrieve`: only the most relevant chunks

Turns the question and every chunk into *embeddings* (lists of numbers that
place similar text close together) and sends only the `top_k` chunks closest to
the question, in a single request. The cost of the answer does not grow with the
document, which makes it the usual choice for finding one fact in something long.
The risk: a chunk that answers the question but does not *look* like it is never
sent.

**Use it for** a specific fact in a long document. **Recipe:** `"needle"`.

### `rerank`: shortlist, then let the model judge

Makes a quick word-matching shortlist (`rerank_candidates` chunks, 20 by default),
asks the model to score each for relevance from 0 to 10, and answers from the
chunks scoring at least `rerank_min_score`. The model's judgement is better than
word matching, at the price of one scoring request per shortlisted chunk.
`skim_model` can do the scoring more cheaply.

If no chunk could be scored, for example because the provider does not support
the structured replies the scoring needs, it falls back to the word-matching
order and warns. If only some chunks were scored, the answer is marked partial.

**Use it for** long documents where word matching alone picks badly, including
scanned ones. **Recipe:** `"scanned"`.

### `iterative`: look, decide what is missing, look again

Retrieves a few chunks and asks the model whether it can answer yet, and if not,
what it should look for next. Then it retrieves again, up to `max_rounds` times. It
suits questions needing several facts from different places ("how did the drop in
margin affect the dividend?"). `ans$notes$queries` records what it searched for in
each round, and `ans$notes$stop_reason` why it stopped.

**Use it for** multi-step questions. **Recipe:** `"research"`.

### `preview`: plan, then read

Shows the model an outline of the document (section headings, sizes, and short
excerpts) and asks which sections to read in full, which to skim, and which to
skip. Then it reads accordingly. A section the plan does not mention is read, never
skipped, so silence cannot lose part of the document; the plan is recorded in
`ans$notes$plan`, so "sections 3 and 5 were not read" is visible. If no usable
plan comes back, it reads everything and marks the answer partial.

**Use it for** long documents with clear sections, when most of them are
irrelevant to the question.

### `ensemble`: several readers, compared

Runs two or more different readers (`members`; by default `retrieve` and
`map_reduce`) and asks the model to reconcile their answers. It costs the sum of
its members plus one request, and pays off when a wrong answer is expensive.
Members must have different signatures, because running the same method twice is
not a second opinion.

**Use it for** high-stakes questions. **Recipe:** `"consensus"`.

### `extract` and `screen`

`extract` fills in a table of details (a *schema*) from every chunk, and `screen`
decides whether a document meets a set of criteria. Both belong to the
literature-review workflow, shown in `vignette("tour")`.

## Choosing a reader

| your situation | reader | cost grows with document length? |
|---|---|---|
| the document fits comfortably in one request | `stuff` | yes, though it is one request |
| one specific fact, long document | `retrieve` | no (embeddings aside) |
| one fact, and word matching picks badly | `rerank` | no, fixed by `rerank_candidates` |
| every mention must be found | `map_reduce` | yes |
| the answer must rest on exact quotations | `skim` | yes |
| an overview of a long document | `hierarchical` | yes |
| an argument that develops through the text | `refine` | yes |
| several facts combined from different places | `iterative` | no, up to `max_rounds` (embeddings aside) |
| long document, most sections irrelevant | `preview` | depends on the plan |
| a wrong answer would be costly | `ensemble` | as its members |

If unsure, run two or three on a document where you already know the answer and
compare (see "Trying readers on your own documents" below).

## Choosing and placing chunks

`retrieve`, `rerank` and `iterative` choose which chunks the model sees. These
settings control the choice and where the chosen chunks go; each says which
readers it applies to.

**`top_k`** (`retrieve`, `rerank`, `iterative`) is how many chunks to send, 6 by
default. For `iterative` it is how many each round adds: the chunks gathered in
earlier rounds stay in the prompt as long as they fit. Too few and the answer
can be missed; too many and you pay for text that does not help.

**`min_score`** (`retrieve`) drops chunks less similar to the question than a
threshold, though the single best chunk is always kept.

**`mmr`** (`retrieve`, `iterative`) stops near-duplicates crowding each other
out. Plain top-k picks the chunks most like the question; if three paragraphs
say the same thing, all three get in. Below 1, `mmr` balances relevance against
similarity to what is already picked (0.7 is a reasonable start):

```{r}
repetitive <- paste(c(
  "Revenue was 45.2 million dollars in fiscal 2024.",
  "Total revenue reached 45.2 million dollars in the 2024 fiscal year.",
  "In fiscal 2024 the company recorded revenue of 45.2 million dollars.",
  "Headcount grew to 1,204 employees across nine clinical sites.",
  "The board approved a dividend of 0.42 dollars per share in March."),
  collapse = "\n\n")
pieces <- gr_segment(gr_ingest(repetitive), list(method = "paragraph", max_tokens = 40))
model <- gr_mock_client(stand_in)
picked <- function(mmr) {
  gr_read(pieces, "What was revenue?", model,
          list(reader = "retrieve", top_k = 3, mmr = mmr))$chunks_used
}
rbind("top-k (mmr = 1)" = picked(1), "mmr = 0.3" = picked(0.3))
```

Plain top-k spent all three places on the same fact; `mmr = 0.3` kept the best
one and used the other two for something different. It costs nothing extra.

**`context_order`** (`retrieve`, `rerank`, `iterative`) decides where the chosen
chunks sit in the prompt. Models attend best to the beginning and end of a long
prompt, so `"edges"` puts the best chunk first and the second-best last;
`"document"` keeps the order they appear in the document, which reads more
naturally when chunks are consecutive; the default `"relevance"` puts the best
first.

**`restate`** (every reader but `extract` and `screen`) repeats the question
before the excerpts as well as after them, because a question asked once, after
several thousand tokens of text, is a long way from the start of the prompt.
`"auto"` does this when the text is long, `"always"` and `"never"` override.

Whether `mmr`, `context_order` or `restate` help depends on your documents and
model; test them rather than assume.

## Embeddings

`retrieve`, `iterative` and the `semantic` way of cutting chunks need
*embeddings*. By default readgpt asks the model provider for them, which costs a
little and needs an internet connection. `gr_options(embedder = "lexical")` uses
word-matching vectors computed locally instead: free, offline and exactly
repeatable, but they match words rather than meaning, so "turnover" will not find
"revenue". Use lexical vectors for practice and testing, and real embeddings for
real work.

```{r}
gr_embedders()[, c("name", "deterministic", "description")]
```

## Asking for citations

With `cite = TRUE` the model is asked to cite the chunks it used, as `[chunk 3]`.
Each chunk id leads back to a page and section through `ans$evidence`:

```{r}
citing <- gr_mock_client(function(messages, params)
  "Revenue was 45.2 million dollars [chunk 2].")
ans <- gr_read(chunks, "What was revenue in 2024?", citing,
               list(reader = "stuff", cite = TRUE))
ans$answer
ans$evidence[ans$evidence$chunk_id == 2, c("chunk_id", "section")]
```

A citation to a chunk the model was never sent is caught and recorded in
`ans$notes$cited_unknown`, and marks the answer partial. `hierarchical` does not
support citations, because it answers from summaries rather than chunks.

## Checking what an answer rests on

`ans$evidence` holds different things depending on the reader:

| reader | `evidence$text` holds |
|---|---|
| `stuff`, `retrieve`, `rerank`, `iterative` | the chunks, word for word |
| `skim` | passages the model copied out |
| `map_reduce` | each chunk's answer |
| `refine`, `hierarchical` | nothing (the answer is built from drafts or summaries) |

Chunks sent word for word are true copies of the document by construction. The
passages `skim` collects are what the model *says* the document contains, so
readgpt checks every one against the chunk it came from:

```{r}
skimmed <- gr_read(chunks, "What was revenue in 2024?", gr_mock_client(stand_in), "skim")
gr_verify_evidence(skimmed)[, c("chunk_id", "verified", "match")]
```

`match` is 1 for an exact quotation (ignoring spacing, capitals and quote marks).
A quotation that is not in the document, such as a changed figure or an invented
sentence, scores lower and is marked `verified = FALSE`. `ans$notes$unverified_evidence`
counts such quotations, and any one of them makes the answer partial.

## Trying readers on your own documents

The best way to choose is to test on a few documents where you already know the
answer. `gr_compare()` runs several configurations on one document, extracting the
text once and sharing chunks between configurations that cut them the same way:

```{r}
cmp <- gr_compare(readgpt_example(), "What was revenue in 2024?", list(
  gr_recipe("narrow", segment = list(method = "structural", max_tokens = 120),
            read = list(reader = "retrieve", top_k = 2)),
  gr_recipe("wide", segment = list(method = "structural", max_tokens = 120),
            read = list(reader = "retrieve", top_k = 6)),
  gr_recipe("all", segment = list(method = "structural", max_tokens = 120),
            read = list(reader = "map_reduce"))),
  client = gr_mock_client(stand_in))
cmp$summary[, c("recipe", "reader", "settings", "chunks_used", "not_found", "partial")]
```

`settings` names whatever each configuration changed from the defaults, so rows
that differ only in a number can be told apart. Two configurations that would do
exactly the same work are run once, with a warning, rather than paid for twice.

## When a provider cannot do structured replies

`rerank`, `iterative` and `preview` ask the model to reply in a fixed structure
(JSON), which most current providers support. Where the structure is not supported
or the reply cannot be read, each one falls back, warns, and marks the answer
partial: `rerank` to word-matching order, `iterative` to a single retrieval,
`preview` to reading everything. The answer is still produced; `partial` tells
you it came by the fallback.

```{r include = FALSE}
gr_options(old)
```

## Next

- `vignette("ingest")`: getting the text out of files in the first place.
- `vignette("tour")`: the ways of cutting documents into chunks, caching and
  replaying runs, reading many documents, and literature reviews.
- `?gr_read_spec`: every reading setting, with its default and allowed range.
